\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[table]{xcolor}
\usepackage{amsmath,amssymb,amsfonts,mathtools,bm}
\usepackage{physics}
\usepackage{booktabs,longtable,array}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage[most]{tcolorbox}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{titlesec}
\usepackage{mathrsfs}

\definecolor{Navy}{HTML}{1F3A5F}
\definecolor{SoftBlue}{HTML}{EAF2F8}
\definecolor{SoftGray}{HTML}{F7F9FA}
\definecolor{RuleGray}{HTML}{D6DBDF}

\pagestyle{fancy}
\fancyhf{}
\lhead{MHD\_PINN\_proj1}
\rhead{Module 2 Physics Notes}
\cfoot{\thepage\ of \pageref{LastPage}}
\setlength{\headheight}{14pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

\titleformat{\section}{\large\bfseries\color{Navy}}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{Navy}}{\thesubsection}{0.6em}{}
\titleformat{\subsubsection}{\normalsize\bfseries}{\thesubsubsection}{0.6em}{}

\newcommand{\uvec}{\mathbf{u}}
\newcommand{\Bvec}{\mathbf{B}}
\newcommand{\Uvec}{\mathbf{U}}
\newcommand{\ptot}{p_{\mathrm{tot}}}
\newcommand{\xhat}{\hat{\mathbf{x}}}
\newcommand{\yhat}{\hat{\mathbf{y}}}
\newcommand{\zhat}{\hat{\mathbf{z}}}

\newtcolorbox{overviewbox}{
  colback=SoftBlue,
  colframe=Navy,
  boxrule=0.7pt,
  arc=2pt,
  left=8pt,right=8pt,top=8pt,bottom=8pt
}
\newtcolorbox{physicsbox}{
  colback=SoftGray,
  colframe=RuleGray,
  boxrule=0.5pt,
  arc=2pt,
  left=8pt,right=8pt,top=8pt,bottom=8pt
}

\begin{document}

\begin{center}
    {\LARGE\bfseries Module 2: Alfv\'en-Wave Verification and Magnetic-Pressure Pulse}\\[0.6em]
    {\large MHD\_PINN\_proj1 Physics Notes}\\[0.4em]
    {\normalsize Revised to be explicit, step-by-step, and derivation-heavy}
\end{center}

\vspace{0.8em}

\begin{overviewbox}
These notes explain the two benchmark problems used after the baseline ideal-MHD equations are in place:
(1) a small-amplitude Alfv\'en wave for \emph{verification}, and
(2) a localized magnetic-pressure pulse for a \emph{nonlinear two-dimensional demonstration}.
The emphasis is not just on what the initial conditions are, but on why those problems were chosen, how their physics follows from the ideal-MHD equations, and what diagnostics genuinely test the solver.
\end{overviewbox}

\section{Symbol and variable table}
\rowcolors{2}{SoftGray}{white}
\begin{longtable}{>{\raggedright\arraybackslash}p{0.18\textwidth} >{\raggedright\arraybackslash}p{0.28\textwidth} >{\raggedright\arraybackslash}p{0.46\textwidth}}
\toprule
\textbf{Symbol} & \textbf{Name} & \textbf{Meaning} \\
\midrule
\endfirsthead
\toprule
\textbf{Symbol} & \textbf{Name} & \textbf{Meaning} \\
\midrule
\endhead
$\rho_0$ & background mass density & Uniform density of the unperturbed equilibrium. \\
$p_0$ & background gas pressure & Uniform thermal pressure of the unperturbed equilibrium. \\
$B_0$ & background magnetic-field magnitude & Constant background field used in the Alfv\'en benchmark. \\
$A$ & perturbation amplitude & Size of the sinusoidal or Gaussian perturbation imposed initially. \\
$k$ & wave number & Spatial frequency of the sinusoidal Alfv\'en mode. \\
$\lambda$ & wavelength & Spatial period of the Alfv\'en wave, $\lambda=2\pi/k$. \\
$v_A$ & Alfv\'en speed & Characteristic propagation speed of the shear Alfv\'en mode. \\
$T$ & wave period & Time required for the wave to travel one wavelength. \\
$u_z$ & out-of-plane velocity & Transverse velocity component used by the benchmark Alfv\'en mode. \\
$B_z$ & out-of-plane magnetic perturbation & Transverse magnetic perturbation coupled to $u_z$. \\
$x_0,y_0$ & pulse-center coordinates & Center of the Gaussian magnetic-pressure pulse. \\
$\sigma$ & pulse width & Characteristic radius of the Gaussian pulse. \\
$\Delta x,\Delta y$ & cell widths & Uniform grid spacing in the two coordinate directions. \\
$\Delta t$ & time-step size & Explicit time step set by the Courant--Friedrichs--Lewy stability restriction. \\
$L^2$ error & root-mean-square error norm & Discrete measure of the mismatch between numerical and reference solutions. \\
$\nabla\cdot\Bvec$ & magnetic divergence & Quantity that must remain small for a physically acceptable MHD computation. \\
\bottomrule
\end{longtable}

\section{Why these two cases are used together}
The project needs more than one test because one benchmark cannot answer every numerical question. The two chosen cases serve different roles:
\begin{enumerate}[leftmargin=1.8em]
    \item The Alfv\'en wave is a \textbf{verification problem}. It has a known propagation speed and a simple analytical solution. That makes it ideal for asking, ``Is the solver reproducing the correct MHD wave physics?''
    \item The magnetic-pressure pulse is a \textbf{nonlinear demonstration problem}. It does not have a simple closed-form exact solution, but it does produce interpretable two-dimensional dynamics driven by magnetic overpressure. That makes it ideal for asking, ``Can the solver evolve a genuinely nonlinear 2D MHD configuration without immediately producing nonsense?''
\end{enumerate}

\begin{physicsbox}
\textbf{Important distinction.} A verification case checks correctness against known mathematics. A demonstration case checks whether the code can produce believable physics in a more complicated setting. The project needs both.
\end{physicsbox}

\section{The Alfv\'en wave from the ideal-MHD equations}

\subsection{Background state}
The benchmark begins from a simple uniform equilibrium:
\begin{equation}
\rho = \rho_0, \qquad
p = p_0, \qquad
\uvec_0 = (0,0,0), \qquad
\Bvec_0 = (B_0,0,0).
\label{eq:m2-bg}
\end{equation}
This means:
\begin{itemize}[leftmargin=1.8em]
    \item the plasma is initially at rest,
    \item density and thermal pressure are constant everywhere,
    \item the magnetic field is straight and points along the $x$-direction.
\end{itemize}
That straight background field is important because the Alfv\'en wave propagates \emph{along} magnetic field lines.

\subsection{What kind of perturbation is added}
A small transverse perturbation is placed in the out-of-plane components:
\begin{equation}
u_z(x,0) = A\sin(kx),
\label{eq:m2-uz-init}
\end{equation}
\begin{equation}
B_z(x,0) = -\sqrt{\rho_0}\,A\sin(kx).
\label{eq:m2-bz-init}
\end{equation}
All other perturbations are zero. This is a very deliberate choice. It means:
\begin{itemize}[leftmargin=1.8em]
    \item the disturbance varies only along the field direction $x$,
    \item the velocity perturbation is perpendicular to $\Bvec_0$,
    \item the magnetic perturbation is also perpendicular to $\Bvec_0$,
    \item the disturbance is small enough that linearization is justified.
\end{itemize}

\subsection{Why the restoring force is magnetic tension}
The momentum equation from Module~1 contains the magnetic-force density in the form
\begin{equation}
\mathbf{f}_B
=
-\nabla\left(\frac{B^2}{2}\right)
+
(\Bvec\cdot\nabla)\Bvec
\label{eq:m2-mag-force}
\end{equation}
in the normalized units used by the project.

For the benchmark equilibrium, the background field is straight and uniform, so the magnetic-pressure gradient of the background is zero. The physically active force is therefore the field-line-tension response to the transverse bend created by the perturbation. The field behaves like a tension-bearing structure: if a field line is bent sideways, the curvature produces a restoring force that tries to straighten it.

\subsection{Derivation of the reduced Alfv\'en system}
We now derive the reduced equations rather than simply stating them.

Start from the ideal-MHD momentum equation in nonconservative form:
\begin{equation}
\rho \frac{D\uvec}{Dt}
=
-\nabla p
-\nabla\left(\frac{B^2}{2}\right)
+
(\Bvec\cdot\nabla)\Bvec.
\label{eq:m2-mom-full}
\end{equation}
Also start from the induction equation:
\begin{equation}
\pdv{\Bvec}{t}
=
\nabla\times(\uvec\times\Bvec).
\label{eq:m2-induction-full}
\end{equation}

Now impose the Alfv\'en-wave assumptions:
\begin{equation}
\rho=\rho_0, \qquad p=p_0, \qquad
\uvec=(0,0,u_z(x,t)), \qquad
\Bvec=(B_0,0,B_z(x,t)).
\label{eq:m2-alfven-assumptions}
\end{equation}
These assumptions say that the disturbance is one-dimensional in $x$, transverse in $z$, and incompressible to leading order.

\subsubsection{Step 1: the $z$-component of the momentum equation}
The left side of \eqref{eq:m2-mom-full} becomes
\begin{equation}
\rho_0 \frac{Du_z}{Dt}.
\end{equation}
Because the perturbation depends only on $x$ and $t$, and because the background flow in $x$ is zero, the material derivative reduces to
\begin{equation}
\frac{Du_z}{Dt} = \pdv{u_z}{t}.
\end{equation}

Now examine the right side. The pressure gradient has no $z$-component because the perturbation is transverse and pressure remains uniform to leading order. The magnetic-pressure term also has no $z$-component in this linearized geometry, because it points along spatial gradients of $B^2$, which are along $x$. The term that survives is the magnetic-tension term:
\begin{equation}
\bigl[(\Bvec\cdot\nabla)\Bvec\bigr]_z.
\end{equation}
Since
\begin{equation}
\Bvec\cdot\nabla = B_0\pdv{}{x} + B_z\pdv{}{z},
\end{equation}
and there is no $z$-dependence, this becomes
\begin{equation}
\bigl[(\Bvec\cdot\nabla)\Bvec\bigr]_z
=
B_0 \pdv{B_z}{x}.
\end{equation}
Therefore the $z$-component of the momentum equation reduces to
\begin{equation}
\rho_0 \pdv{u_z}{t} = B_0 \pdv{B_z}{x}.
\label{eq:m2-uz-eq1}
\end{equation}
Equivalently,
\begin{equation}
\pdv{u_z}{t} - \frac{B_0}{\rho_0}\pdv{B_z}{x}=0.
\label{eq:m2-uz-reduced}
\end{equation}

\subsubsection{Step 2: the $z$-component of the induction equation}
Now compute $\uvec\times\Bvec$:
\begin{equation}
\uvec\times\Bvec
=
\begin{vmatrix}
\xhat & \yhat & \zhat \\
0 & 0 & u_z \\
B_0 & 0 & B_z
\end{vmatrix}
=
(0,\;u_z B_0,\;0).
\label{eq:m2-cross}
\end{equation}
Take the curl:
\begin{equation}
\nabla\times(\uvec\times\Bvec)
=
\nabla\times(0,u_z B_0,0).
\end{equation}
Because the quantity depends only on $x$,
\begin{equation}
\nabla\times(0,u_z B_0,0)
=
\left(0,\;0,\;\pdv{}{x}(u_z B_0)\right).
\end{equation}
Since $B_0$ is constant,
\begin{equation}
\pdv{B_z}{t} = B_0 \pdv{u_z}{x}.
\label{eq:m2-bz-reduced}
\end{equation}

\subsubsection{Step 3: combine the two equations into a wave equation}
Take a time derivative of \eqref{eq:m2-uz-reduced}:
\begin{equation}
\pdv[2]{u_z}{t} - \frac{B_0}{\rho_0}\pdv{}{x}\left(\pdv{B_z}{t}\right)=0.
\end{equation}
Substitute \eqref{eq:m2-bz-reduced}:
\begin{equation}
\pdv[2]{u_z}{t} - \frac{B_0}{\rho_0}\pdv{}{x}\left(B_0 \pdv{u_z}{x}\right)=0.
\end{equation}
Because $B_0$ is constant,
\begin{equation}
\pdv[2]{u_z}{t} - \frac{B_0^2}{\rho_0}\pdv[2]{u_z}{x}=0.
\label{eq:m2-wave-u}
\end{equation}
This is the ordinary wave equation with speed
\begin{equation}
v_A = \frac{B_0}{\sqrt{\rho_0}}.
\label{eq:m2-va}
\end{equation}
The same derivation holds for $B_z$.

\subsection{Exact traveling-wave solution}
For the initial sinusoidal mode, the solution is a translated profile:
\begin{equation}
u_z(x,t)=A\sin\left[k(x-v_A t)\right],
\label{eq:m2-exact-uz}
\end{equation}
\begin{equation}
B_z(x,t)=-\sqrt{\rho_0}\,A\sin\left[k(x-v_A t)\right].
\label{eq:m2-exact-bz}
\end{equation}
The wavelength and period are
\begin{equation}
\lambda = \frac{2\pi}{k},
\qquad
T = \frac{\lambda}{v_A}.
\label{eq:m2-lambda-period}
\end{equation}

\subsection{Why the solver can be judged quantitatively}
Because the benchmark has an explicit analytical solution, the solver can be tested with concrete diagnostics.

\subsubsection{$L^2$ error}
For a discrete line cut $q_i^{\text{num}}$ and exact values $q_i^{\text{exact}}$, the discrete root-mean-square error is
\begin{equation}
\|e\|_{L^2}
=
\sqrt{
\frac{1}{N_x}
\sum_{i=1}^{N_x}
\left(q_i^{\text{num}}-q_i^{\text{exact}}\right)^2 }.
\label{eq:m2-l2}
\end{equation}
A smaller error means the waveform shape and amplitude are being captured more accurately.

\subsubsection{Measured phase speed}
The code also measures how far the waveform has shifted in time and compares that against the predicted speed $v_A$. This is important because a numerical method could, in principle, preserve shape reasonably well but move the wave at the wrong speed. The benchmark therefore tests both \emph{shape fidelity} and \emph{characteristic-speed fidelity}.

\subsubsection{Why the final time matters}
If the benchmark is run for exactly one full wave period, a periodic sinusoid returns to the same phase modulo the domain length. Then a crude shift-based diagnostic can report zero net displacement even though the wave propagated correctly. For that reason, measuring the phase shift at a quarter period or another nondegenerate fraction of the period is often a better diagnostic choice.

\section{Why the solver is upgraded from first order to second order}

\subsection{What first-order diffusion does to a smooth wave}
A first-order piecewise-constant finite-volume scheme with a diffusive numerical flux tends to smear smooth structures. For the Alfv\'en benchmark, that means:
\begin{itemize}[leftmargin=1.8em]
    \item the sinusoidal crest becomes too rounded,
    \item the trough becomes too shallow,
    \item the amplitude decays more than it should,
    \item the line-cut comparison against the exact solution becomes visibly blurred.
\end{itemize}
That is not a failure of the underlying physics model; it is a limitation of the numerical approximation.

\subsection{Why piecewise-linear reconstruction helps}
Instead of treating every cell as having a constant state, the piecewise-linear method (PLM) reconstructs a linear profile within each cell:
\begin{equation}
q(x) \approx q_i + \sigma_i (x-x_i),
\label{eq:m2-plm}
\end{equation}
where $q_i$ is the cell average and $\sigma_i$ is a slope estimate. The reconstructed left and right interface states are
\begin{equation}
q_{i+1/2}^{L}=q_i+\frac12 \sigma_i \Delta x,
\qquad
q_{i+1/2}^{R}=q_{i+1}-\frac12 \sigma_{i+1}\Delta x.
\label{eq:m2-plm-interfaces}
\end{equation}
This makes the interface data much more faithful to a smooth sinusoidal field.

\subsection{Why a limiter is still necessary}
If the slopes are taken naively, oscillations can appear near sharp gradients. The minmod limiter suppresses that by selecting the smaller of the forward and backward slopes when they have the same sign, and setting the slope to zero when they do not:
\begin{equation}
\operatorname{minmod}(a,b)
=
\begin{cases}
\operatorname{sign}(a)\min(|a|,|b|), & ab>0,\\
0, & ab\le 0.
\end{cases}
\label{eq:m2-minmod}
\end{equation}
So the method becomes second-order accurate in smooth regions but retreats toward first order in dangerous regions.

\subsection{Why Runge--Kutta 2 is paired with PLM}
The spatial approximation alone is not the whole story. If the time update remains only first-order accurate, some of the benefit of the improved spatial reconstruction is lost. The two-stage Runge--Kutta method therefore updates the state via
\begin{equation}
\Uvec^{(1)} = \Uvec^n + \Delta t\,\mathcal{L}(\Uvec^n),
\end{equation}
\begin{equation}
\Uvec^{n+1}
=
\frac12
\left[
\Uvec^n
+
\Uvec^{(1)}
+
\Delta t\,\mathcal{L}(\Uvec^{(1)})
\right],
\label{eq:m2-rk2}
\end{equation}
where $\mathcal{L}$ is the finite-volume spatial operator. The physical solution is unchanged; what improves is the fidelity with which the numerical method follows the continuous evolution.

\section{The magnetic-pressure pulse as a nonlinear demonstration}

\subsection{Why magnetic pressure can launch motion}
From Module~1, the magnetic force can be written as
\begin{equation}
\mathbf{f}_B
=
-\nabla\left(\frac{B^2}{2}\right)
+
(\Bvec\cdot\nabla)\Bvec.
\end{equation}
The first term behaves like a pressure gradient built from magnetic energy density. Therefore, if one region has a larger $B^2$ than its surroundings, that region is \emph{magnetically overpressured} and pushes outward on the surrounding fluid.

\subsection{How the pulse is initialized}
The demonstration case introduces a localized out-of-plane magnetic perturbation:
\begin{equation}
B_z(x,y,0)
=
A\exp\left[
-\frac{(x-x_0)^2+(y-y_0)^2}{2\sigma^2}
\right].
\label{eq:m2-gaussian}
\end{equation}
The density, thermal pressure, and in-plane magnetic field are otherwise taken uniform. The Gaussian profile means the field is strongest near $(x_0,y_0)$ and decreases smoothly away from the center.

\subsection{Why this creates a magnetic-pressure gradient}
Because $B_z$ is largest at the pulse center, the magnetic energy density
\begin{equation}
e_{\mathrm{mag}} = \frac12 |\Bvec|^2
\end{equation}
is also largest there. Therefore the magnetic-pressure contribution
\begin{equation}
p_{\mathrm{mag}}=\frac12 |\Bvec|^2
\end{equation}
has a spatial gradient pointing outward from the pulse center. The force density contribution
\begin{equation}
-\nabla p_{\mathrm{mag}}
\end{equation}
therefore pushes fluid away from the region of strongest field.

\subsection{Why the pulse is nonlinear}
Unlike the Alfv\'en verification problem, the Gaussian magnetic-pressure pulse is not just a translated copy of itself. As it evolves,
\begin{itemize}[leftmargin=1.8em]
    \item the field redistributes,
    \item velocity develops in multiple directions,
    \item density and gas pressure can change,
    \item nonlinear couplings between field and flow become important.
\end{itemize}
So this case asks the solver to do more than follow a single linear wave family. It asks the solver to handle a genuinely two-dimensional MHD response.

\begin{physicsbox}
\textbf{Interpretation.} The pulse is not mainly about having an exact formula. It is about checking whether the code can produce a physically interpretable nonlinear MHD evolution without obvious pathologies such as exploding divergence error, negative pressure, or wildly asymmetric artifacts unrelated to the problem setup.
\end{physicsbox}

\subsection{Why the diagnostics change for this case}
Since there is no simple pointwise exact solution, the diagnostic focus changes:
\begin{enumerate}[leftmargin=1.8em]
    \item monitor total mass conservation,
    \item monitor total-energy behavior,
    \item monitor $\max |\nabla\cdot\Bvec|$,
    \item inspect maps of density, pressure, velocity magnitude, and magnetic-field magnitude.
\end{enumerate}
These diagnostics do not prove exact correctness the way the Alfv\'en benchmark does, but they do check whether the nonlinear evolution remains physically credible.

\section{Why magnetic divergence must remain part of the workflow}
Both benchmark problems must satisfy
\begin{equation}
\nabla\cdot\Bvec = 0.
\label{eq:m2-divb}
\end{equation}
Even if the solver does not yet enforce this constraint exactly with a specialized method such as constrained transport, it must still monitor the quantity. Large divergence error can feed back into the momentum equation and create unphysical forces. That is why the divergence diagnostic is not optional bookkeeping; it is part of the physics validation logic.

\section{What Module 2 accomplishes}
After Module~1 introduces the governing equations, Module~2 turns them into a testing strategy:
\begin{itemize}[leftmargin=1.8em]
    \item the Alfv\'en wave checks whether the solver reproduces a known MHD propagation problem with the correct speed and low numerical error,
    \item the magnetic-pressure pulse checks whether the same solver can evolve a nonlinear and spatially two-dimensional magnetic overpressure problem in a believable way,
    \item the second-order upgrade is motivated by the need to capture the smooth benchmark more faithfully while keeping the method robust near stronger gradients.
\end{itemize}
These ideas form the bridge between the basic equations and the more advanced numerical and PINN modules that follow.

\end{document}
