\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage[table]{xcolor}
\usepackage{amsmath,amssymb,amsfonts,mathtools,bm}
\usepackage{physics}
\usepackage{booktabs,longtable,array}
\usepackage{enumitem}
\usepackage[hidelinks]{hyperref}
\usepackage[most]{tcolorbox}
\usepackage{fancyhdr}
\usepackage{lastpage}
\usepackage{titlesec}

\definecolor{Navy}{HTML}{1F3A5F}
\definecolor{SoftBlue}{HTML}{EAF2F8}
\definecolor{SoftGray}{HTML}{F7F9FA}
\definecolor{RuleGray}{HTML}{D6DBDF}

\pagestyle{fancy}
\fancyhf{}
\lhead{MHD\_PINN\_proj1}
\rhead{Module 4 Physics Notes}
\cfoot{\thepage\ of \pageref{LastPage}}
\setlength{\headheight}{14pt}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

\titleformat{\section}{\large\bfseries\color{Navy}}{\thesection}{0.6em}{}
\titleformat{\subsection}{\normalsize\bfseries\color{Navy}}{\thesubsection}{0.6em}{}
\titleformat{\subsubsection}{\normalsize\bfseries}{\thesubsubsection}{0.6em}{}

\newcommand{\Uvec}{\mathbf{U}}
\newcommand{\Fvec}{\mathbf{F}}
\newcommand{\Gvec}{\mathbf{G}}
\newcommand{\Iten}{\mathbf{I}}
\newcommand{\uvec}{\mathbf{u}}
\newcommand{\Bvec}{\mathbf{B}}
\newcommand{\ptot}{p_{\mathrm{tot}}}

\newtcolorbox{overviewbox}{
  colback=SoftBlue,
  colframe=Navy,
  boxrule=0.7pt,
  arc=2pt,
  left=8pt,right=8pt,top=8pt,bottom=8pt
}
\newtcolorbox{physicsbox}{
  colback=SoftGray,
  colframe=RuleGray,
  boxrule=0.5pt,
  arc=2pt,
  left=8pt,right=8pt,top=8pt,bottom=8pt
}

\begin{document}

\begin{center}
    {\LARGE\bfseries Module 4: Numerical Method for the Conventional MHD Solver}\\[0.6em]
    {\large MHD\_PINN\_proj1 Physics Notes}\\[0.4em]
    {\normalsize Revised to connect every algorithmic piece to the underlying conservation laws}
\end{center}

\vspace{0.8em}

\begin{overviewbox}
These notes explain how the conventional solver translates the ideal-MHD equations into a computable algorithm.
The emphasis is not merely on listing the numerical ingredients. The goal is to show why a finite-volume method is the natural choice for conservation laws, how interface fluxes arise, why reconstruction and time integration are upgraded, and how each numerical decision relates to the underlying physics.
\end{overviewbox}

\section{Symbol and variable table}
\rowcolors{2}{SoftGray}{white}
\begin{longtable}{>{\raggedright\arraybackslash}p{0.19\textwidth} >{\raggedright\arraybackslash}p{0.27\textwidth} >{\raggedright\arraybackslash}p{0.46\textwidth}}
\toprule
\textbf{Symbol} & \textbf{Name} & \textbf{Meaning} \\
\midrule
\endfirsthead
\toprule
\textbf{Symbol} & \textbf{Name} & \textbf{Meaning} \\
\midrule
\endhead
$N_x,N_y$ & grid cell counts & Number of finite-volume cells in the two coordinate directions. \\
$\Delta x,\Delta y$ & cell widths & Uniform spacing of the Cartesian grid. \\
$\Delta t$ & time step & Explicit update step chosen from a Courant stability bound. \\
$\Uvec_{i,j}$ & cell-averaged conserved state & Average conserved state stored in cell $(i,j)$. \\
$\Fvec_{i+1/2,j}$ & numerical flux through an $x$-face & Flux leaving or entering a cell across a vertical face. \\
$\Gvec_{i,j+1/2}$ & numerical flux through a $y$-face & Flux leaving or entering a cell across a horizontal face. \\
$a_{\max}$ & maximum signal speed & Largest estimated wave speed relevant to a face flux. \\
$c_{f,x},c_{f,y}$ & fast magnetosonic speeds & Fastest characteristic speeds normal to the coordinate directions. \\
CFL & Courant number & Dimensionless safety factor controlling the explicit time step. \\
$\Uvec_L,\Uvec_R$ & left and right interface states & Reconstructed states used to solve a face Riemann problem approximately. \\
$\sigma_i$ & reconstructed slope & Cellwise slope used by piecewise-linear reconstruction. \\
$\operatorname{minmod}$ & minmod limiter & Nonlinear limiter used to reduce spurious oscillations near steep gradients. \\
$\mathcal{L}(\Uvec)$ & spatial operator & Finite-volume flux-difference operator appearing in the semidiscrete system. \\
$\nabla\cdot\Bvec$ & magnetic divergence & Diagnostic quantity that should remain small in a valid MHD solution. \\
\bottomrule
\end{longtable}

\section{Why the project begins with a conventional solver}
Before a physics-informed neural network can be judged honestly, the project needs a direct numerical method that solves the intended equations in a transparent way. The conventional solver supplies:
\begin{itemize}[leftmargin=1.8em]
    \item a baseline for what the equations predict,
    \item physically meaningful diagnostics,
    \item a way to distinguish modeling error from neural-network training error,
    \item a source of reference solutions for later comparison.
\end{itemize}

\begin{physicsbox}
\textbf{Scientific reason.} A PINN without a conventional baseline is hard to interpret. If the network prediction looks plausible, there is no trustworthy anchor. If it looks implausible, there is no clear way to separate physics error from optimizer failure.
\end{physicsbox}

\section{Why a finite-volume method is the natural language of conservation laws}

\subsection{Start from the conservative PDE}
The ideal-MHD system is written in conservation-law form:
\begin{equation}
\pdv{\Uvec}{t}
+
\pdv{\Fvec(\Uvec)}{x}
+
\pdv{\Gvec(\Uvec)}{y}
=
0.
\label{eq:m4-pde}
\end{equation}
This equation says that the time rate of change of a conserved quantity inside a region is determined by the net flux through the region's boundary.

\subsection{Integrate over one control volume}
Take one grid cell:
\begin{equation}
V_{ij}=[x_{i-1/2},x_{i+1/2}] \times [y_{j-1/2},y_{j+1/2}].
\end{equation}
Integrate \eqref{eq:m4-pde} over that cell:
\begin{equation}
\int_{V_{ij}}
\pdv{\Uvec}{t}\,dA
+
\int_{V_{ij}}
\pdv{\Fvec}{x}\,dA
+
\int_{V_{ij}}
\pdv{\Gvec}{y}\,dA
=
0.
\label{eq:m4-integral1}
\end{equation}
Bring the time derivative outside the integral:
\begin{equation}
\dv{}{t}
\int_{V_{ij}} \Uvec\,dA
+
\int_{V_{ij}}
\pdv{\Fvec}{x}\,dA
+
\int_{V_{ij}}
\pdv{\Gvec}{y}\,dA
=
0.
\label{eq:m4-integral2}
\end{equation}

\subsection{Convert flux-divergence terms to face terms}
Because the grid is Cartesian,
\begin{equation}
\int_{V_{ij}}\pdv{\Fvec}{x}\,dA
=
\int_{y_{j-1/2}}^{y_{j+1/2}}
\left[
\Fvec(x_{i+1/2},y)-\Fvec(x_{i-1/2},y)
\right]dy.
\end{equation}
Likewise,
\begin{equation}
\int_{V_{ij}}\pdv{\Gvec}{y}\,dA
=
\int_{x_{i-1/2}}^{x_{i+1/2}}
\left[
\Gvec(x,y_{j+1/2})-\Gvec(x,y_{j-1/2})
\right]dx.
\end{equation}
Define the cell average
\begin{equation}
\Uvec_{ij}
=
\frac{1}{\Delta x \Delta y}
\int_{V_{ij}} \Uvec\,dA.
\label{eq:m4-cellavg}
\end{equation}
Then the exact finite-volume balance law becomes
\begin{equation}
\dv{\Uvec_{ij}}{t}
=
-\frac{1}{\Delta x}
\left(
\Fvec_{i+1/2,j}-\Fvec_{i-1/2,j}
\right)
-\frac{1}{\Delta y}
\left(
\Gvec_{i,j+1/2}-\Gvec_{i,j-1/2}
\right).
\label{eq:m4-semidiscrete}
\end{equation}

\subsection{What this means physically}
Equation \eqref{eq:m4-semidiscrete} has a simple interpretation:
\begin{itemize}[leftmargin=1.8em]
    \item if more mass leaves a cell than enters, the cell mass decreases,
    \item if more momentum leaves than enters, the cell momentum decreases,
    \item if more total energy leaves than enters, the cell energy decreases.
\end{itemize}
The same face-flux logic applies to each row of the conserved vector.

\section{Why a numerical flux is needed}
At a cell face, the left and right neighboring cells generally do not agree. So the physical face flux is ambiguous: which state should be used? That is the local Riemann-problem idea. The project begins with a robust approximate answer: the Rusanov flux.

\subsection{Rusanov flux}
At a face with left and right states $\Uvec_L$ and $\Uvec_R$,
\begin{equation}
\Fvec^{\mathrm{num}}
=
\frac12\left[\Fvec(\Uvec_L)+\Fvec(\Uvec_R)\right]
-
\frac12 a_{\max}(\Uvec_R-\Uvec_L).
\label{eq:m4-rusanov}
\end{equation}
The same structure applies in the $y$ direction with $\Gvec$.

\subsection{Interpretation of the two terms}
The first term is a centered average of the physical fluxes. If the left and right states are similar, this part dominates. The second term is dissipative. It damps sharp jumps in proportion to a wave-speed estimate.

That second term is why Rusanov is robust:
\begin{itemize}[leftmargin=1.8em]
    \item if a discontinuity tries to create unstable numerical behavior, the dissipative term opposes it,
    \item the dissipation is stronger when the characteristic speeds are larger,
    \item the price paid is extra numerical diffusion.
\end{itemize}

\subsection{Why the signal speed must be large enough}
The speed $a_{\max}$ must dominate the fastest information that can cross the face. In MHD, a safe choice is the fast magnetosonic speed. For propagation normal to the $x$ direction,
\begin{equation}
c_{f,x}^{2}
=
\frac12\left[
a^{2}+b^{2}
+
\sqrt{(a^{2}+b^{2})^{2}-4a^{2}b_x^{2}}
\right],
\label{eq:m4-cfx}
\end{equation}
where
\begin{equation}
a^{2}=\frac{\gamma p}{\rho},
\qquad
b^{2}=\frac{B_x^{2}+B_y^{2}+B_z^{2}}{\rho},
\qquad
b_x^{2}=\frac{B_x^{2}}{\rho}.
\label{eq:m4-soundmag}
\end{equation}
Then a common estimate is
\begin{equation}
a_{\max,x}=|u_x|+c_{f,x}.
\label{eq:m4-amaxx}
\end{equation}
The same logic produces $a_{\max,y}=|u_y|+c_{f,y}$.

\subsection{Why fast magnetosonic speed is the conservative choice}
MHD has several wave families. The fast mode is the broadest characteristic envelope, so using it is a conservative stability decision. It may add more diffusion than necessary in some places, but it protects the method from underestimating how fast information can move.

\section{Why the time step must satisfy a CFL condition}

\subsection{Physical intuition}
An explicit method updates the solution using only information from the current time level. Therefore one time step cannot be allowed to move information across too many cells. Otherwise the update would be trying to propagate signals farther than the stencil can represent consistently.

\subsection{CFL form}
A common stability restriction is
\begin{equation}
\Delta t
=
\mathrm{CFL}
\cdot
\min
\left(
\frac{\Delta x}{a_{\max,x}},
\frac{\Delta y}{a_{\max,y}}
\right).
\label{eq:m4-cfl}
\end{equation}
If the characteristic speeds grow, the allowed time step shrinks. This is not arbitrary. It reflects the physical fact that faster wave motion must be resolved by smaller explicit time increments.

\section{Why piecewise-linear reconstruction is needed}

\subsection{What goes wrong with piecewise-constant data}
If each cell is treated as constant, then every face sees left and right states that jump sharply from one average to the next. That is robust, but it smears smooth features. The Alfv\'en benchmark makes this obvious because a smooth sinusoid becomes too diffused.

\subsection{Piecewise-linear reconstruction}
In one dimension, approximate a quantity $q$ inside cell $i$ by
\begin{equation}
q(x)\approx q_i+\sigma_i(x-x_i),
\label{eq:m4-linear}
\end{equation}
where $\sigma_i$ is a slope estimate. Then the left and right edge states are
\begin{equation}
q_i^{L}=q_i-\frac12 \sigma_i \Delta x,
\qquad
q_i^{R}=q_i+\frac12 \sigma_i \Delta x.
\label{eq:m4-lr}
\end{equation}
This gives the numerical flux a more realistic estimate of what the solution looks like at the interface.

\subsection{Why a limiter is needed}
A naive centered slope may create new extrema or oscillations near steep gradients. The minmod limiter avoids that by comparing one-sided differences. If
\begin{equation}
a=\frac{q_i-q_{i-1}}{\Delta x},
\qquad
b=\frac{q_{i+1}-q_i}{\Delta x},
\end{equation}
then
\begin{equation}
\operatorname{minmod}(a,b)
=
\begin{cases}
\operatorname{sign}(a)\min(|a|,|b|), & ab>0,\\
0, & ab\le 0.
\end{cases}
\label{eq:m4-minmod}
\end{equation}
So:
\begin{itemize}[leftmargin=1.8em]
    \item in a smooth monotone region, the slope is retained in a cautious way,
    \item near a local extremum or sign change, the slope is suppressed,
    \item the method becomes less oscillatory and more robust.
\end{itemize}

\section{Why Runge--Kutta 2 is used for time integration}
The semidiscrete system \eqref{eq:m4-semidiscrete} can be written abstractly as
\begin{equation}
\dv{\Uvec}{t}=\mathcal{L}(\Uvec),
\label{eq:m4-ode}
\end{equation}
where $\mathcal{L}$ is the flux-difference operator. A first-order Euler step is
\begin{equation}
\Uvec^{n+1}=\Uvec^n+\Delta t\,\mathcal{L}(\Uvec^n),
\end{equation}
but that is only first-order accurate in time.

The two-stage Runge--Kutta update is
\begin{equation}
\Uvec^{(1)}=\Uvec^n+\Delta t\,\mathcal{L}(\Uvec^n),
\end{equation}
\begin{equation}
\Uvec^{n+1}
=
\frac12\Uvec^n
+
\frac12\left[
\Uvec^{(1)}+\Delta t\,\mathcal{L}(\Uvec^{(1)})
\right].
\label{eq:m4-rk2}
\end{equation}
This gives a better time approximation for smooth evolution. Since PLM is already improving the spatial approximation, pairing it with a second-order time integrator makes the overall method more balanced.

\section{Why boundary conditions matter physically}
For the Alfv\'en verification problem, periodic boundaries are used in $x$. That means:
\begin{equation}
\Uvec(x_{\min},y,t)=\Uvec(x_{\max},y,t).
\end{equation}
Physically, this corresponds to a repeating domain. Numerically, it allows a traveling sinusoid to leave one side and re-enter the other without artificial reflection.

In the code, periodicity is often implemented through ghost cells copied from the opposite side. That sounds like a code detail, but it is actually a direct numerical representation of the domain topology.

\section{Why magnetic divergence is monitored as a diagnostic}
The exact equations require
\begin{equation}
\nabla\cdot\Bvec=0.
\label{eq:m4-divb}
\end{equation}
In two dimensions,
\begin{equation}
\nabla\cdot\Bvec
=
\pdv{B_x}{x}+\pdv{B_y}{y}.
\end{equation}
A finite-volume code that does not use a specialized divergence-control method will generally not preserve this quantity exactly. Monitoring it is therefore essential. If the diagnostic grows too large, the numerical solution may develop unphysical magnetic forces.

\section{Why positivity floors are included}
Density and gas pressure must remain physically meaningful:
\begin{equation}
\rho>0,\qquad p>0.
\end{equation}
Near strong gradients or under-resolved features, an explicit scheme can sometimes produce small negative values due to numerical error. A positivity floor is a pragmatic safeguard:
\begin{itemize}[leftmargin=1.8em]
    \item it prevents the update from entering a mathematically invalid state,
    \item it keeps wave-speed calculations and flux functions defined,
    \item it makes debugging easier during development.
\end{itemize}
It is not the final ideal cure, but it is an honest and useful development tool.

\section{How one full solver step fits together}
Putting all the pieces together, one upgraded solver step looks like this:
\begin{enumerate}[leftmargin=1.8em]
    \item start from cell-averaged conserved variables,
    \item recover primitive variables where needed,
    \item reconstruct left and right interface states with PLM and minmod limiting,
    \item compute numerical fluxes with the Rusanov formula,
    \item assemble the finite-volume spatial operator $\mathcal{L}$,
    \item advance with Runge--Kutta 2,
    \item optionally apply positivity floors,
    \item evaluate diagnostics such as $L^2$ error and $\max|\nabla\cdot\Bvec|$.
\end{enumerate}
Each of those steps traces back to a physical or mathematical requirement:
conservation, characteristic propagation, stability, accuracy, or admissibility.

\begin{physicsbox}
\textbf{Bottom line.} The conventional solver is not a random collection of MATLAB routines. It is a structured translation of the conservation-law form of ideal MHD into a sequence of operations that preserve flux balance, approximate wave propagation, control numerical instability, and expose meaningful diagnostics.
\end{physicsbox}

\section{What Module 4 accomplishes}
Module~4 explains why the conventional solver is built the way it is:
\begin{itemize}[leftmargin=1.8em]
    \item finite volume because the equations are conservation laws,
    \item face fluxes because conserved quantities change by crossing cell boundaries,
    \item Rusanov because it is robust and easy to implement,
    \item PLM plus minmod because the solver must improve smooth-wave resolution without becoming oscillatory,
    \item Runge--Kutta 2 because the time integration should match the upgraded spatial accuracy,
    \item diagnostics and floors because development must remain physically honest.
\end{itemize}
That makes the numerical method understandable as physics translated into an algorithm, not merely as a list of coding choices.

\end{document}
